---
output:
  rmarkdown::github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

```{r, include = FALSE}
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
}
```

# `atmle`: Adaptive TMLE for RCT + RWD Data Fusion

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

This package implements adaptive targeted minimum loss-based estimation of
average treatment effects using combined randomized trial and real-world data to
improve efficiency.

__Authors:__ [Sky Qiu](https://github.com/tq21),
Jens Tarp, 
[Lars van der Laan](https://larsvanderlaan.github.io/),
[Mark van der Laan](https://vanderlaan-lab.org/),

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/tq21/atmle/issues).

---

## Example

```{r atmle-ate-fusion-example, message=FALSE, warning=FALSE, fig.width=8, fig.height=4}
library(atmle)
library(sl3)
library(ggplot2)

set.seed(2026)
n <- 600
data <- data.frame(W1 = rbinom(n, 1, .5), W2 = rbinom(n, 1, .5),
                   S = rbinom(n, 1, .5), A = rbinom(n, 1, .5))
data$Y <- with(data, rbinom(n, 1,
  plogis(-1 + .4*W1 + .3*W2 + .5*A + .3*(1-S)*A)))
W_nodes <- c("W1", "W2")
cells <- expand.grid(W1 = 0:1, W2 = 0:1)
true_ate <- with(cells, mean(plogis(-1+.4*W1+.3*W2+.5) -
                             plogis(-1+.4*W1+.3*W2)))

# unadjusted trial-only difference-in-means
rct <- data[data$S == 1, ]
diff_est <- with(rct, mean(Y[A == 1])-mean(Y[A == 0]))
diff_se <- sqrt(with(rct,
                     var(Y[A == 1])/sum(A == 1)+var(Y[A == 0])/sum(A == 0)))
diff_ci <- diff_est + c(-1, 1) * qnorm(0.975) * diff_se

# A-TMLE
lrnr <- list(
  learners = list(sl3::Lrnr_glm$new())
)
fit <- atmle_ate_fusion$new(
  data = data,
  S_node = "S",
  W_nodes = W_nodes,
  A_node = "A",
  Y_node = "Y",
  family = "binomial",
  n_folds = 3
)

fit$run(
  g_bar_method = lrnr,
  theta_method = lrnr,
  Pi_method = lrnr,
  Q_bar_method = lrnr,
  A_cate_args = list(max_degree = 3L, smoothness_orders = 1L, num_knots = 5L),
  S_cate_args = list(max_degree = 3L, smoothness_orders = 1L, num_knots = 5L,
                     force_A = TRUE),
  target_method = "tmle",
  target_gwt = FALSE,
  max_iter = 100,
  verbose = FALSE
)
```

```{r results-table}
fit$results[, c("param", "estimator", "psi", "se", "lower", "upper",
                 "unforced_weight")]
```

Each population has its original A-TMLE row plus one variance-floor convex
combination row. `se` is the influence-curve standard error, and `lower` and
`upper` are the corresponding Wald limits. The following plot displays these
intervals.

```{r}
atmle_res <- fit$results[fit$results$param == "Avg. over S=1", ]
df_plot <- data.frame(Estimator = c("Unadjusted Difference-in-Means",
                                    "A-TMLE", "Convex combination"),
                      Estimate = c(diff_est, atmle_res$psi),
                      Lower = c(diff_ci[1], atmle_res$lower),
                      Upper = c(diff_ci[2], atmle_res$upper))
df_plot$Estimator <- factor(df_plot$Estimator, 
                            levels = c("Unadjusted Difference-in-Means",
                                       "A-TMLE", "Convex combination"))
ggplot(df_plot, aes(x = Estimator, y = Estimate, fill = Estimator)) +
  geom_point() +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2, linewidth = 0.8) +
  geom_hline(yintercept = true_ate, linetype = "dashed", color = "red") +
  theme_minimal(base_size = 13) +
  labs(title = "Point Estimates with 95% Confidence Intervals",
       y = "Estimate",
       x = "") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  coord_flip()
```

## Convex combination and influence-curve inference

The package fits both unforced and forced-A source-bias HAL models using common
nuisance estimates and CV folds. `force_A` selects which original A-TMLE rows to
display; the combination is always added. For unforced estimate U and forced
estimate F, define Vu, Vf, C, and Vd from their paired influence curves: the two
sample variances, covariance, and variance of the difference, each divided by n.
The combination is `F + w*(U-F)`, where

```r
bias_squared <- max((U - F)^2, Vd)
w <- min(weight_cap, max(0, (Vf - C) / (Vd + bias_squared)))
```

The default `weight_cap = 1` allows the full convex range, with the unforced
weight between 0 and 1. A smaller cap can be supplied explicitly. Degenerate
difference variances give weight zero. Weights are computed separately for the
two population targets.
When `n_lambda > 1`, all original working-model rows remain, and the combination
uses each candidate's first (CV-selected) working model.

The combined influence curve is `(1-w)*D_F + w*D_U`. The estimated variance is
`var((1-w)*D_F + w*D_U)/n`; its square root is the reported standard error `se`.
Wald limits use this standard error and a normal critical value. This
calculation treats the estimated weight as fixed and does not include
HAL-selection uncertainty.

`fit$components` contains both candidates' results and paired influence curves;
`fit$influence` is aligned with the rows of `fit$results`.
`fit$inference(alpha = 0.1)` recalculates the Wald intervals at 90% confidence
from the existing fit.


## License

&copy; 2026 [Sky Qiu](https://github.com/tq21), 
Jens Tarp, 
[Lars van der Laan](https://larsvanderlaan.github.io/),
[Mark van der Laan](https://vanderlaan-lab.org/),

The contents of this repository are distributed under the MIT license.
