---
title: "A-TMLE Package Tutorial: Basic Functionalities"
author: "Sky Qiu"
date: "2024-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
In this vignette, we will demonstrate an application of the A-TMLE R package
for estimating the average treatment effect in a setting where we would like to
integrate data from both a randomized controlled trial (RCT) and real-world data
(RWD). The `atmle` R package implements the adaptive targeted minimum loss-based
estimation (A-TMLE) for estimating the average treatment effect (ATE) in the 
data fusion/data integration setting.

## Installation
In this vignette, we will be using the `atmle` package to construct an A-TMLE
for the ATE integrating both RCT and RWD. We will also use the `tmle` package
to construct a regular TMLE on the RCT data as a comparison. In addition, we will 
use the `ggplot2` package for visualization.
Note that the `atmle` and `sl3` packages are not on CRAN yet, so you need to 
install them from GitHub. The code to install the packages is provided in the code 
chunk below.

```{r, message=FALSE, warning=FALSE}
#install.packages("remotes", "origami", "purrr", "data.table", "tmle", "ggplot2")
#remotes::install_github("tq21/atmle")
#remotes::install_github("tlverse/sl3")
#remotes::install_github("tq21/hal9001@kmeans-knot-screen")
library(atmle)
library(tmle)
library(ggplot2)
```

## Data Generating Process
Suppose we have conducted a randomized trial where we would like to evaluate
the effect of some hypertension medication on patients' systolic blood pressure.
We have the following baseline covariates measured:
- Age: patient age (mean = 50, sd = 12);
- BMI: patient body mass index (mean = 28, sd = 5);
- Smoking: patient smoking status (20% smokers in the population).
Suppose the true effect of the medication on systolic blood pressure is -10mmHg.
That is, if the patient took the medication, their systolic blood pressure would
be 10mmHg lower than if they did not take the medication.

In addition to the RCT, we also have access to some RWD where the treatment is
assigned based on patient characteristics. In particular, we have two unmeasured 
confounders that affect both the treatment assignment and systolic blood pressure:
health literacy and socioeconomic status.

In the next code chunk, we have a function `sim_data()` to generate the combined 
data set (~30% RCT, ~70% RWD).

```{r}
sim_data <- function(n) {
  pct_rct <- 0.3 # the proportion of RCT in the total sample
  n_rct <- round(n * pct_rct)
  n_rwd <- n - n_rct

  # covariates for all participants
  Age <- round(rnorm(n, mean = 50, sd = 12))
  BMI <- round(rnorm(n, mean = 28, sd = 5), 2)
  Smoking <- rbinom(n, 1, 0.2)

  # unmeasured confounders for RWD
  Health_Literacy <- rnorm(n, mean = 50, sd = 10)
  Socioeconomic_Status <- rnorm(n, mean = 50, sd = 10)

  # treatment assignment
  # RCT: randomized treatment
  A_rct <- rbinom(n_rct, 1, 0.5)

  # RWD: treatment assignment based on observed and unobserved covariates
  logit_p_rwd <- -1.5 + 0.05 * Age[(n_rct+1):n] +
    0.04 * BMI[(n_rct+1):n] +
    0.8 * Smoking[(n_rct+1):n] +
    0.03 * Health_Literacy[(n_rct+1):n] + 0.02 * Socioeconomic_Status[(n_rct+1):n]
  prob_treatment_rwd <- plogis(logit_p_rwd)
  A_rwd <- rbinom(n_rwd, 1, prob_treatment_rwd)

  A <- c(A_rct, A_rwd)

  # outcome regression with non-linearity
  base_effect <- 120 - 0.1 * Age + 0.05 * BMI^2 - 5 * Smoking +
    0.04 * Health_Literacy + 0.03 * Socioeconomic_Status
  treatment_effect <- -10 * A
  random_effect <- rnorm(n)
  Y <- base_effect + treatment_effect + random_effect

  # study indicator
  S <- c(rep(1, n_rct), rep(0, n_rwd))

  return(data.frame(Age = Age,
                    BMI = BMI,
                    Smoking = Smoking,
                    A = A,
                    Y = round(Y, 2),
                    S = S))
}
```

We can simulate 1000 patients from the DGP. Make sure that the data has study 
indicator $S$, baseline covariates $W$, treatment $A$, and outcome $Y$.
```{r}
set.seed(128947)
data <- sim_data(n = 1000)
head(data)
```

Now, we will estimate the ATE using A-TMLE. We need to specify the column indices
for the nodes and the learners for the nuisance parameters and working models.
We also need to set the `controls_only` argument to `FALSE`, because we have both
treatment and control arm in the RWD. For estimating the nuisance parameters,
we will use `sl3`, the super learner package in `tlverse`, to estimate the nuisance 
parameters. The default learner library can be found in `R/get_default_sl3_learners.R` 
script. One could also specify their own learner library following the documentation of `atmle()` function.
We will use highly adaptive lasso (HAL) to learn the bias working model and the
conditional treatment effect working model.
```{r}
atmle_res <- atmle(data = data,
                   S_node = 6,
                   W_node = c(1, 2, 3),
                   A_node = 4,
                   Y_node = 5,
                   controls_only = FALSE,
                   family = "gaussian",
                   theta_method = "sl3",
                   Pi_method = "sl3",
                   g_method = "sl3",
                   theta_tilde_method = "sl3",
                   bias_working_model = "HAL",
                   pooled_working_model = "HAL",
                   g_rct = 0.5,
                   enumerate_basis_args = list(num_knots = 50),
                   verbose = TRUE)
```

Now, let's estimate the ATE using a regular TMLE on the RCT data only.
```{r}
# TMLE on the RCT data only
data_rct <- data[data$S == 1,]
tmle_res <- tmle(
  Y = data_rct$Y, A = data_rct$A, W = data_rct[, c("Age", "BMI", "Smoking")], 
  g1W = rep(0.5, nrow(data_rct)),
  Q.SL.library = c("SL.glm"),
  family = "gaussian"
)
print(tmle_res)
```

Now, we will visually compare the results of the two estimators using the `ggplot2` package.
Note that because of how we defined our target parameter, we need to add a minus
sign to the estimates if we would like to interpret them as reductions in systolic blood pressure.
The true average treatment effect is marked with a red dashed line.
```{r}
# plot results
res_df <- data.frame(
  Estimator = c("RCT-only", "A-TMLE"),
  est = c(-tmle_res$estimates$ATE$psi, -atmle_res$est),
  lower = c(-tmle_res$estimates$ATE$CI[1], -atmle_res$lower),
  upper = c(-tmle_res$estimates$ATE$CI[2], -atmle_res$upper))

plt <- ggplot(res_df, aes(x = est, y = Estimator)) +
  geom_point(position = position_dodge(width = 0.5), size = 3.5) +
  geom_errorbar(aes(xmin = lower, xmax = upper),
                position = position_dodge(width = 0.5), width = 0.3, linewidth = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 1.5) +
  geom_vline(xintercept = 10, linetype = "dashed", linewidth = 1.5, color = "red") +
  labs(title = "Estimated Reduction in Systolic Blood Pressure",
       x = "", y = "", color = "") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title.x = element_text(size = 14, face = "bold"),
        axis.title.y = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12),
        legend.text = element_text(size = 14))
plt
```

From the plot, we can see that the 95% confidence interval produced by A-TMLE 
is much narrower than that of the RCT-only TMLE estimator.
