---
title: "A-TMLE Package Tutorial: An Example of Data Fusion"
author: "Sky Qiu"
date: "2024-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
In this vignette, we will demonstrate an application of the A-TMLE R package
for estimating the average treatment effect in a setting where data from both an
RCT and real-world are available.

## Data Generating Process
In this section, we write a function `sim_data()` to generate an RCT, in which
the treatment assignment is randomized, and real-world data, in which the
treatment assignment depends on patient characteristics. We will mimic the
realistic scenario where the real-world data is subject to unmeasured
confounding.

Suppose we have conducted a randomized trial where we would like to evaluate
the effect of some hypertension medication on patients' systolic blood pressure.
We have the following baseline covariates:
- Age: patient age (mean = 50, sd = 12);
- BMI: patient body mass index (mean = 28, sd = 5);
- Smoking: patient smoking status (20% smokers in the population).
Suppose the true effect of the medication on systolic blood pressure is -10mmHg.
That is, if the patient took the medication, their systolic blood pressure would
be 10mmHg lower than if they did not take the medication.

- `A`: treatment assignment
- `Y`: outcome
- `Age`: age
- `BMI`: body mass index
- `Smoking`: smoking status

TODO: write down the SCM

```{r}
sim_data <- function(n) {
  pct_rct <- 0.3 # the proportion of RCT in the total sample
  n_rct <- round(n * pct_rct)
  n_rwd <- n - n_rct
  
  # covariates for all participants
  Age <- rnorm(n, mean=50, sd=12) # Mean age of 50 with sd of 12
  BMI <- rnorm(n, mean=28, sd=5)  # Mean BMI of 28 with sd of 5
  Smoking <- rbinom(n, 1, 0.2)    # 20% smokers in the population
  
  # unmeasured confounders for RWD
  Health_Literacy <- rnorm(n_rwd, mean=50, sd=10)
  Socioeconomic_Status <- rnorm(n_rwd, mean=50, sd=10)

  # treatment assignment
  # RCT: randomized treatment
  A_rct <- rbinom(n_rct, 1, 0.5)
  
  # RWD: treatment assignment based on observed and unobserved covariates
  logit_p_rwd <- -1 + 0.05 * Age[(n_rct+1):n] + 
    0.04 * BMI[(n_rct+1):n] + 
    0.8 * Smoking[(n_rct+1):n] + 
    0.03 * Health_Literacy + 0.02 * Socioeconomic_Status
  prob_treatment_rwd <- plogis(logit_p_rwd)
  A_rwd <- rbinom(n_rwd, 1, prob_treatment_rwd)
  
  A <- c(A_rct, A_rwd)
  
  # Outcome model with non-linear and interaction effects
  base_effect <- 120 - 0.1 * Age + 0.05 * BMI^2 - 5 * Smoking
  treatment_effect <- -10 * A
  random_effect <- rnorm(n)
  Y <- base_effect + treatment_effect + random_effect
  
  # Add effects of unmeasured confounders to RWD outcomes
  Y[(n_rct+1):n] <- Y[(n_rct+1):n] + 0.04 * Health_Literacy + 0.03 * Socioeconomic_Status
  
  # Study indicator
  S <- c(rep(1, n_rct), rep(0, n_rwd))
  
  # Assemble the dataframe without the unmeasured confounders (as they are not observed)
  data.frame(Age = Age, BMI = BMI, Smoking = Smoking, A = A, Y = Y, S = S)
}

# Example usage:
set.seed(123) # For reproducibility
simulated_data <- sim_data(1000)
head(simulated_data)

```

```{r}
library(atmle)
`%+%` <- function(a, b) paste0(a, b)
```

```{r}
set.seed(143)
data <- sim_data(3000)
atmle_res <- atmle(data = data,
                   S_node = 6,
                   W_node = c(1, 2, 3),
                   A_node = 4,
                   Y_node = 5,
                   controls_only = FALSE,
                   family = "gaussian",
                   atmle_pooled = TRUE,
                   theta_method = "glm",
                   Pi_method = "glm",
                   g_method = "glm",
                   g_delta_method = "glm",
                   theta_tilde_method = "glm",
                   bias_working_model = "glmnet",
                   pooled_working_model = "glmnet",
                   g_rct = 0.5,
                   var_method = "ic",
                   v_folds = 5,
                   g_bounds = c(0.01, 0.99),
                   Pi_bounds = c(0.01, 0.99),
                   theta_bounds = c(-Inf, Inf),
                   target_gwt = TRUE)
```


```{r}
atmle_res$lower
atmle_res$upper
```

```{r}
library(EScvtmle)
escvtmle_res <- ES.cvtmle(txinrwd = TRUE,
                          data = data,
                          study = "S",
                          covariates = c("Age", "BMI", "Smoking"),
                          treatment_var = "A",
                          treatment = 1,
                          outcome = "Y",
                          pRCT = 0.5,
                          family = "gaussian",
                          Q.SL.library = c("SL.glm"),
                          g.SL.library = c("SL.glm"),
                          Q.discreteSL = TRUE,
                          g.discreteSL = TRUE,
                          V = 5)
```


```{r}
atmle_res$lower
atmle_res$upper
```

```{r}
escvtmle_res
```


